Symbolic Weighted Language Models and\\ Parsing over Infinite Alphabets

### 2021-05-23 main2.tex
revision algo best search for new extended def. VPA

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA) 
over  $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$, $\Semiring$ and $\bar\Phi$ 
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states, 
$P$ is a finite set of stack symbols, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar\wei$ is a sextuplet composed of the transition functions:
$\weii : Q \times P \times Q \to \Phici$,  
$\weiei : Q \times Q \to \Phii$,  
$\weic : Q \times P \times Q \times P \to \Phicc$,  
$\weiec : Q \times P \times Q \to \Phic$,  
$\weir : Q \times P \times Q \to \Phicr$,  
$\weier : Q \times Q \to \Phir$.
%and 
%$\weiex : Q \times Q \to \Phix$ 
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer}, 
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$, 
$a \in \Omegai$, 
$\call{c} \in \Omegac$, 
$\return{r} \in \Omegar$, 
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times \Omegac \times P \times \Omegai \times Q \to \Semiring & 
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) & 
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Omegai \times Q \to \Semiring & 
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\
%
\weic: Q \times \Omegac \times P \times  \Omegac \times P \times Q \to \Semiring & 
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) & 
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times \Omegac \times P \times Q \to \Semiring & 
\weiec(q, {\call{c}}, p, q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\
%
\weir: Q \times \Omegac \times P \times \Omegar \times Q \to \Semiring & 
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) & 
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Omegar \times Q \to \Semiring & 
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}      
\]

@@

\noindent
$\weii$ and $\weiei$ both read an input internal symbol $a$ and change state from $q$ to $q'$, 
without changing the stack. 
Moreover, $\weii$ reads 
${\call{c}} \in \Omegac$ and $p \in P$ at the top of the stack 
($c$ is compared to $a$ by the weight function $\eta_\mathsf{ci} \in \Phici$)
and $\weiei$ applies when the stack is empty.

\noindent
$\weic$ and $\weiec$ read the input call symbol $\call{c}'$, 
push it to the stack along with $p'$, and change state from $q$ to to $q'$.
Moreover, $\weic$ reads ${\call{c}}$ and $p$ at the top of the stack 
($c$ is compared to $c'$),
and $\weiec$ applies when the stack is empty.

\noindent
$\weir$ and $\weier$ read the input return symbol $\return{r}$, and change state from $q$ to to $q'$.
Moreover, $\weir$ reads and pop from stack a pair made of $\call{c}$ and $p$, 
($\call{c}$ is compared to $\return{r}$),
and $\weier$ applies when the stack is empty.
%In this case, the weight function $\phi_\mathsf{r}$ 
%computes a value of matching between the call and return symbols $c$ and $r$.
%This value might be set to $\zero$ in order to express that the symbols do not match.

@@

\begin{example}
structured words...
intro language of music notation ?
\end{example}

\subsection{Properties}
Like VPA and symbolic VPA, 
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$ and $\Semiring$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as in the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.

\subsection{Best Search} 
\label{sec:best}\label{sec:search}
**hypotheses**
Let us assume that the semiring~$\Semiring$ is
commutative, bounded, complete, and total.
and assume an effective label theory.
%s.t. for all $\phi \in \Phii$, 
%$\psi \in \Phir$, 
%and $\eta \in \Phicr$, 
%$\displaystyle\bigoplus_{a \in \Omegai} \phi(a)$ 
%$\displaystyle\bigoplus_{r \in \Omegar} \phi(r)$ and 
%$\displaystyle\bigoplus_{{\call{c}} \in \Omegac} 
%\displaystyle\bigoplus_{{\return{r}} \in \Omegar} \eta({\call{c}}, {\return{r}})$
%are computable...

@@

\noindent
More precisely, 
let $\top$ be a fresh stack symbol which does not belong to $\Gamma$,
and for every two states $q, q' \in Q$ and $\sigma \in \{ \bot, \top \}$, 
let 
\[ 
  d_0(q, \sigma, q') = \bigoplus_{s\in \Omega^*} 
  \textstyle
  \weight_A\bigl(\config{q}{\sigma}, s, \config{q'}{\sigma}\bigr). 
\]
Since $\Semiring$ is complete, this infinite sum is well defined, 
and since $\leq_\oplus$ is assumed total, it is the minimum in $\Omega^*$
of the fonction 
$s \mapsto \weight_A(\config{q}{\sigma}, s, \config{q'}{\sigma})$ 
\wrt this ordering.
%
When $\sigma = \bot$, $d_0(q, \sigma, q')$ is the central expression in %a term of 
the definition (\ref{eq:weightA}) of $A(s_0)$ for the minimum $s_0$
(for the function $\weight_A$).
%
When $\sigma = \top$, intuitively, it is the minimum weight of a computation of $A$
starting in state $q$ with a stack $\gamma \in \Gamma^*$ (possibly empty), 
and ending in state $q'$ with the same stack $\gamma$, such that moreover 
the computation does not touch a symbol of $\gamma$. 
That means that during the computation, 
$A$ may apply the first case of in the definition of $\weight_A$ (internal symbol),
as well as the second case, to push call symbols on the top of $\gamma$, 
and may pop these symbols with the third case (return symbol). 
However, it cannot apply one of the two last cases (return symbol and empty stack)
when the current stack is $\gamma$.
%pop symbols in $\gamma$.
% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).
%
Algorithm~\ref{algo:Dijkstra}
constructs iteratively a marking $d : Q \times \{ \bot, \top \} \times Q \to \Semiring$
%of the triplets $\<q, \sigma, q'>$ 
%of states of $A$ by weight values in $\Semiring$, 
that converges eventually to $d_0$.
%It uses for that purpose a priority queue $P$ containing triplets of 
%$Q \times \{ \bot, \top \} \times Q$.

@@

\begin{figure}[t]
\[
\begin{array}{rrcl}
\multicolumn{4}{l}{\mathrm{For~all~} q_0, q_3 \in Q, p \in P,}\\
 & d(q_1, \sigma, q_3) & \opluseq &
   d(q_1, \sigma, q_2) \otimes \displaystyle\bigoplus_{a \in \Omegai} \weii(q_2, a, q_3)\\
%
%  & d(q_0, \sigma, q_2) & \opluseq & 
%    \displaystyle\bigoplus_{a \in \Omegai} \weii(q_0, a, q_1) \otimes d(q_1, \sigma, q_2)\\
%
\mathrm{if~} \sigma = \top & d(q_0, \top, q_3) & \opluseq &
d(q_1, \sigma, q_2) \otimes
%\displaystyle\bigoplus_{p \in P} 
\displaystyle\bigoplus_{{\call{c}} \in \Omegac} 
\displaystyle\bigoplus_{{\return{r}} \in \Omegar} \eta({\call{c}}, {\return{r}})\\
\mathrm{and} &  d(q_0, \bot, q_3) & \opluseq &
d(q_1, \sigma, q_2) \otimes
%\displaystyle\bigoplus_{p \in P} 
\displaystyle\bigoplus_{{\call{c}} \in \Omegac} 
\displaystyle\bigoplus_{{\return{r}} \in \Omegar} \eta({\call{c}}, {\return{r}})\\
& & & \mathrm{where~} \eta = \weic(q_0, q_1, p) \otimes \weir(q_2, p, q_3)\\
%
\mathrm{if~} \sigma = \bot & d(q_1, \bot, q_3) & \opluseq &
  d(q_1, \sigma, q_2) \otimes \displaystyle\bigoplus_{\return{r} \in \Omegar} \weie(q_2, {\return{r}}, q_3)\\
%
%& d(q_0, \bot, q_2) & \opluseq & 
%d(q_0, \bot, q_1) \otimes d(q_1, \sigma, q_2), \mathrm{if~} \< q_0, \bot, q_1> \notin P \\
%
%\mathrm{if~} \sigma = \top & d(q_0, \top, q_2) & \opluseq & 
%d(q_0, \top, q_1) \otimes d(q_1, \sigma, q_2), \mathrm{if~} \< q_0, \top, q_1> \notin P \\
%
& d(q_1, \bot, q_3) & \opluseq & 
   d(q_1, \sigma, q_2) \otimes d(q_2, \bot, q_3), \mathrm{if~} \< q_2, \bot, q_3> \notin P \\
%
\mathrm{if~} \sigma = \top & d(q_1, \top, q_3) & \opluseq & 
    d(q_1, \sigma, q_2) \otimes d(q_2, \top, q_3), \mathrm{if~} \< q_2, \top, q_3> \notin P \\
\end{array}
\]
\caption{Update $d$ with $d(q_1, \sigma, q_2)$.}
\label{fig:best-update}
\end{figure}

>>>>

\begin{definition}
A \emph{Symbolic Weighted Visibly Pushdown Automata} (\SWVPA) 
over  $\Omega = \Omegai \uplus \Omegac \uplus \Omegar$, $\Semiring$ and $\bar\Phi$ 
is a tuple $A = \< Q, P, \init, \bar\wei, \final >$,
where $Q$ is a finite set of states, 
$P$ is a finite set of stack symbols, 
$\mathsf{in} : Q \to \Semiring$, 
respectively $\mathsf{out} : Q \to \Semiring,$
are functions defining the weight for entering, 
respectively leaving, a state, 
and $\bar\wei$ is a sextuplet composed of the transition functions:
$\weii : Q \times P \times Q \to \Phici$,  
$\weiei : Q \times Q \to \Phii$,  
$\weic : Q \times P \times Q \times P \to \Phicc$,  
$\weiec : Q \times P \times Q \to \Phic$,  
$\weir : Q \times P \times Q \to \Phicr$,  
$\weier : Q \times Q \to \Phir$.
%and 
%$\weiex : Q \times Q \to \Phix$ 
%with $\mathsf{x} \in \{ \mathsf{i}, \mathsf{c}, \mathsf{r}\}$.
\end{definition}
%
Similarly as in Section~\ref{section:transducer}, 
we extend the above transition functions as follows
for all $q, q' \in Q$, $p \in P$, 
$a \in \Omegai$, 
$\call{c} \in \Omegac$, 
$\return{r} \in \Omegar$, 
overloading their names: % for simplicity:
\[
\begin{array}{lll}
\weii: Q \times \Omegac \times P \times \Omegai \times Q \to \Semiring & 
\weii(q, c, p, a, q') = \eta_\mathsf{ci}(c, a) & 
\mathrm{where~} \eta_\mathsf{ci} = \weii(q, p, q'),\\
%
\weiei: Q \times \Omegai \times Q \to \Semiring & 
\weiei(q, a, q') = \phi_\mathsf{i}(a) &
\mathrm{where~} \phi_\mathsf{i} = \weiei(q, q').\\[2pt]
%
\weic: Q \times \Omegac \times P \times  \Omegac \times P \times Q \to \Semiring & 
\weic(q, c, p, \call{c'}, p', q') = \eta_\mathsf{cc}(c, \call{c'}) & 
\mathrm{where~} \eta_\mathsf{cc} = \weic(q, p, p', q'),\\
%
\weiec: Q \times \Omegac \times P \times Q \to \Semiring & 
\weiec(q, {\call{c}}, p, q') = \phi_\mathsf{c}({\call{c}}) &
\mathrm{where~} \phi_\mathsf{c} = \weiec(q, p, q').\\[2pt]
%
\weir: Q \times \Omegac \times P \times \Omegar \times Q \to \Semiring & 
\weir(q, {\call{c}},  p, {\return{r}}, q') = \eta_\mathsf{cr}({\call{c}},  {\return{r}}) & 
\mathrm{where~} \eta_\mathsf{cr} = \weir(q, p, q'),\\
%
\weier: Q \times \Omegar \times Q \to \Semiring & 
\weier(q, {\return{r}}, q') = \phi_\mathsf{r}({\return{r}}) &
\mathrm{where~} \phi_\mathsf{r} = \weier(q, q').\\
\end{array}      
\]

@@

\noindent
$\weii$ and $\weiei$ both read an input internal symbol $a$ and change state from $q$ to $q'$, 
without changing the stack. 
Moreover, $\weii$ reads a pair made of 
${\call{c}} \in \Omegac$ and $p \in P$ at the top of the stack 
($c$ is compared to $a$ by the weight function $\eta_\mathsf{ci} \in \Phici$)
and $\weiei$ applies if and only if the stack is empty.

\noindent
$\weic$ and $\weiec$ read the input call symbol $\call{c}'$, 
push it to the stack along with $p'$, and change state from $q$ to to $q'$.
Moreover, $\weic$ reads ${\call{c}}$ and $p$ at the top of the stack 
($c$ is compared to $c'$),
and $\weiec$ applies iff the stack is empty.

\noindent
$\weir$ and $\weier$ read the input return symbol $\return{r}$, and change state from $q$ to to $q'$.
Moreover, $\weir$ reads and pop from stack a pair made of $\call{c}$ and $p$, 
($\call{c}$ is compared to $\return{r}$),
and $\weier$ applies iff the stack is empty.
%In this case, the weight function $\phi_\mathsf{r}$ 
%computes a value of matching between the call and return symbols $c$ and $r$.
%This value might be set to $\zero$ in order to express that the symbols do not match.

@@

\begin{example}
structured words with timed symbols...
intro language of music notation? (markup = time division, leaves = events etc)
\end{example}

\subsection{Properties}
Like VPA and symbolic VPA, 
the class of \SWVPA is closed under the binary operators of the underlying semiring.
%
\begin{proposition}
Let $A_1$ and $A_2$ be two (\SWVPA)
over the same $\Omega$, $\Semiring$ and $\bar\Phi$.
There exists two $\SWVPA$ $A_1 \oplus A_2$ and $A_1 \otimes A_2$, 
effectively constructible, 
such that for all $s \in \Omega^*$, 
$(A_1 \oplus A_2)(s) = A_1(s) \oplus A_2(s)$ and 
$(A_1 \otimes A_2)(s) = A_1(s) \otimes A_2(s)$.
\end{proposition}
The construction is essentially the same 
as in the case of the Boolean semiring~\cite{dAntonyAlur14SVPDA}.

\subsection{Best Search} 
\label{sec:best}\label{sec:search}
%**hypotheses**
Let us assume that the semiring~$\Semiring$ is
commutative, bounded, complete, and total.
and assume an effective label theory.
%s.t. for all $\phi \in \Phii$, 
%$\psi \in \Phir$, 
%and $\eta \in \Phicr$, 
%$\displaystyle\bigoplus_{a \in \Omegai} \phi(a)$ 
%$\displaystyle\bigoplus_{r \in \Omegar} \phi(r)$ and 
%$\displaystyle\bigoplus_{{\call{c}} \in \Omegac} 
%\displaystyle\bigoplus_{{\return{r}} \in \Omegar} \eta({\call{c}}, {\return{r}})$
%are computable...

@@

\noindent
More precisely, 
let $b_\bot : Q \times Q \to \Semiring$ be the function:
%
\begin{equation}
  b_\bot(q, q') = \bigoplus_{s\in \Omega^*} 
  \textstyle
  \weight_A\bigl(\config{q}{\bot}, s, \config{q'}{\bot}\bigr)
\end{equation}
%
Since $\Semiring$ is complete, this infinite sum is well defined, 
and since $\leq_\oplus$ is assumed total, it is the minimum in $\Omega^*$
of the fonction 
$s \mapsto \weight_A(\config{q}{\sigma}, s, \config{q'}{\sigma})$ 
\wrt this ordering.
%
Let $\top$ be a fresh stack symbol which does not belong to $\Gamma$.
and let $b_\top : Q \times P \times Q \to \Phic$ be such that,
for every two states $q, q' \in Q$ 
and stack symbol $p \in P$ %and $\sigma \in \{ \bot, \top \}$, 
\begin{equation}
  b_\top(q, p, q') : c \mapsto \bigoplus_{s\in \Omega^*} 
  \textstyle
  \weight_A\bigl(\config{q}{c\,p\cdot \top }, s, \config{q'}{c\, p\cdot \top}\bigr). 
\end{equation}
%
When $\sigma = \bot$, $b_\bot(q, q')$ is the central expression in %a term of 
the definition (\ref{eq:weightA}) of $A(s_0)$ for the minimum $s_0$
(for the function $\weight_A$).
%
When $\sigma = \top$, intuitively, it is the minimum weight of a computation of $A$
starting in state $q$ with a stack $\gamma \in \Gamma^*$ (possibly empty), 
and ending in state $q'$ with the same stack $\gamma$, such that moreover 
the computation does not touch a symbol of $\gamma$. 
That means that during the computation, 
$A$ may apply the first case of in the definition of $\weight_A$ (internal symbol),
as well as the second case, to push call symbols on the top of $\gamma$, 
and may pop these symbols with the third case (return symbol). 
However, it cannot apply one of the two last cases (return symbol and empty stack)
when the current stack is $\gamma$.
%pop symbols in $\gamma$.
% Note that having a stack reduced to such a symbol makes impossible the application of the 
% two last cases in the definition of $\weight_A$ (return symbol and empty stack). 
% However, it is possible to apply the two first cases 
% (internal symbol or call symbol, with a push on the top of $\top$).
%
Algorithm~\ref{algo:Dijkstra}
constructs iteratively a marking $d : Q \times \{ \bot, \top \} \times Q \to \Semiring$
%of the triplets $\<q, \sigma, q'>$ 
%of states of $A$ by weight values in $\Semiring$, 
that converges eventually to $d_0$.
%It uses for that purpose a priority queue $P$ containing triplets of 
%$Q \times \{ \bot, \top \} \times Q$.

@@

\begin{figure}
For all $q_0, q_3 \in Q$, %$p \in P$,
\[
\begin{array}{lcl}
%\multicolumn{3}{l}{\mathrm{For~all~} q_0, q_3 \in Q, p \in P,}\\
d_\top(q_1, p, q_3) & \opluseq &
  d_\top(q_1, p, q_2) \otimes 
  \displaystyle\bigoplus_{a \in \Omegai} \weii(q_2, p, q_3)_a\\
\multicolumn{3}{l}{\quad
\mathrm{where~} \weii(q_2, p, q_3)_a \in \Phic
\mathrm{~is~the~partial~application~}
x_c \mapsto \weii(q_2, x_c, p, a, q_3)}\\
%
d_\bot(q_1, p, q_3) & \opluseq &
  d_\bot(q_1, q_2) \otimes 
  \displaystyle\bigoplus_{a \in \Omegai} \weie(q_2, a, q_3)\\
%     
d_\top(q_0, p, q_3) & \opluseq &
  \displaystyle\bigoplus_{{\call{c}'} \in \Omegac} 
  \displaystyle\bigoplus_{{\return{r}} \in \Omegar} 
  \weic(q_0, p, p', q_1)_{c'} \otimes 
  d_\top(q_1, p', q_2) \otimes
  \weir(q_2, c', p', r, q_3)\\
\multicolumn{3}{l}{\quad
\mathrm{where~} \weic(q_0, p, p', q_1)_{c'} \in \Phic
\mathrm{~is~the~partial~application~}
x_c \mapsto \weic(q_0, p, x_c, c', p', q_1)}\\[2pt]
%
d_\bot(q_0, q_3) & \opluseq &
  \displaystyle\bigoplus_{{\call{c}} \in \Omegac} 
  \displaystyle\bigoplus_{{\return{r}} \in \Omegar} 
  \weiec(q_0, c, p, q_1) \otimes 
   d_\top(q_1, p, q_2)(c) \otimes
  \weir(q_2, c, p, r, q_3)\\
%
d_\bot(q_1, q_3) & \opluseq &
  d_\bot(q_1, q_2) \otimes 
  \displaystyle\bigoplus_{\return{r} \in \Omegar} \weier(q_2, {\return{r}}, q_3)\\
%
d_\top(q_1, p, q_3) & \opluseq & 
  d_\top(q_1, p, q_2) \otimes d_\top(q_2, p, q_3), 
  \mathrm{if~} \< q_2, \top, q_3> \notin P \\
%
d_\bot(q_1, q_3) & \opluseq & 
  d_\bot(q_1, q_2) \otimes d_\bot(q_2, q_3), \mathrm{if~} \< q_2, \bot, q_3> \notin P \\
\end{array}
\]
%$\weii(q_2, p, q_3)_a \in \Phic$, for $a \in \Omegai$,
%is the partial application
%$x_c \mapsto \weii(q_2, x_c, p, a, q_3)$.
%
%$\weic(q_0, p, p', q_1)_{c'} \in \Phic$, for $c' \in \Omegac$,
%is the partial application
%$x_c \mapsto \weic(q_0, p, x_c, c', p', q_1)$.
\caption{Update 
$d_\top$ with $d_\top(q_1, p, q_2)$ or $d_\bot$ with $d_\top(\bot, q_2)$.}
\label{fig:best-update}
\end{figure}
